Network Deliberation: The role of network structure in large-scale, internet-enabled, participatory decision-making

### 2022-08-09 UMich/chapters/ch01-intro.tex
Dissertation revisions to abstract and ch01.

The ability of large groups to reach mutually agreeable decisions is key to
democratic governance, organizational effectiveness, social movements, and peer production.
Faced with the intractability of large-scale decision-making, traditional
systems have sacrificed one or more desirable properties, such as participation
(as in representative democracy), deliberation (as in voting), equality
(as in command hierarchies), and speed (as in formal consensus).
This dissertation examines the emerging role of internet-enabled collaborative
networks in overcoming these historical limitations.

In recent years, examples of large-scale collaborations have emerged that seem
to achieve the previously unachievable. Millions of volunteer Wikipedia editors
have created a high-quality encyclopedia without centralized leadership
\cite{keegan_evolution_2017, giles_internet_2005}.
The Free Software movement has produced the Linux kernel and GNU operating
system, which power much of the modern internet
\cite{coleman_coding_2012, benkler_coases_2002, raymond_cathedral_1999}.
Social movements such as the Arab Spring, Occupy, Black Lives Matter, and
Podemos have reshaped national governments and brought attention to deeply
entrenched social issues
\cite{tufekci_twitter_2017, gonzalez-bailon_networked_2016}.
Participatory governance at such scales is unprecedented.
The emergence of these large-scale decentralized collaborations has been
attributed to the fast, bidirectional,
and global communication enabled by the internet
\cite{tufekci_twitter_2017, benkler_coases_2002}.
A better understanding of how specifically such communication sidesteps
historical barriers to large-scale collaboration will contribute to more
effective policy as well as best practices for organizational design and
intervention.
This dissertation focuses on one particularly challenging aspect of
such collaborations: decision-making.
I specificaly examine how,
when groups are too large for all members to participate in all discussions,
the course and outcome of the decision-making process is influenced by the
communication network structure: the shape of who talks to whom.

In this dissertation, I combine observational, agent-based modeling, and experimental methods to evaluate the role of network structure in large-scale participatory decision-making.
Taken together, the findings in these studies suggests that small, tightly-knit pods, interlocked through the sharing of members can benefit such collaborations.
This common structure, which I refer to as network deliberation, appears to preserve some benefits of small-scale deliberation, while allowing efficient diffusion of information through very large groups.
Furthermore, I find evidence that network deliberation can mitigate some of the negative aspects of social influence.
These findings provide insight into the success of existing large-scale collaborations as well as provide guidance for the design of new sociotechnical systems.

@@

The fundamental challenge of large-scale collective decision-making is how to
reconcile the conflicting preferences of individual group members.
This challenge has been studied formally in social choice theory,
a sub-field of economics.
Prior work in social choice theory has found, somewhat discouragingly,
that even when all individual preferences are known perfectly,
making a fair collective decision isn't always possible.
In some cases, the method of aggregating individual preferences
(i.e. the voting system) can influence the outcome.
Social choice theory focuses on understanding of these limitations,
such as the Condorcet Paradox
\cite{condorcet_essay_1785} and
Arrow's Impossibility Theorem \cite{arrow_social_2012}.
This dissertation finds hope in the transformative potential of the
deliberative process.
Social choice theory typically assumes fixed individual preferences.
Deliberation allows individuals to influence and change each other's
preferences,
which creates the potential to sidestep the historical limitations of
social choice theory.
When individual preferences are allowed to vary, it becomes possible for an
irreconcilable set of preferences to evolve into one with a clear winner.
So far, this possibility has received relatively little attention,
most likely due to the historical intractability of large-scale deliberative
decision-making.
This dissertation explores the potential of effective large-scale deliberation, as enabled by the internet.
Such large-scale deliberation creates the potential for members to
resolve conflicting preferences and reach mutually acceptable decisions
without relying on coercive or hierarchical processes that might introduce
power imbalances or informational biases.

@@

Throughout this dissertation,
I motivate and develop a novel theoretical framework,
which I call {\em network deliberation}.
Network deliberation is an analytical framework that can be applied both to the empirical observation of existing sociotechical systems,
as well as to guide the creation and analysis of simplified models.
My review of the literature identifies a common theme
among successful large-scale internet-enabled collaborations:
large collectives composed of interlocking smaller groups.
These groups have various names, including:
committees, working groups, teams, circles, cores, syndicates,
affinity groups, zones, and nodes.
As an abstraction of these small interlocking group, I will use the term ``pods.''
Network deliberation describes large-scale collective deliberation achieved
through interlocking pods.
As in the theories of interlocking directorates \cite{levine_study_1979},
interlocking publics \cite{habermas_structural_1991},
and network rotation \cite{salehi_hive_2018},
pods allow for beneficial small group dynamics,
while the overlap between pods enables diffusion of information and opinions
through the greater collective.
In network deliberation, the method of assigning individuals to pods
(whether deliberate or self-organized) can produce interpersonal networks with
varying structures.
The central questions of this dissertation are 1. does an interlocking-pod structure improve deliberation relative to conventional structures? and 2. how does the structure of these interlocking-pod networks influence the process and outcome of deliberation
in large collaborations?

@@

\section{Contributions}
This dissertation describes the contributions of three projects.
Chapter \ref{chap:wp-prod-perf} describes an observational and computational
study of WikiProjects on the English-language Wikipedia,
and reports the following contributions:
\begin{itemize}
\setlength\itemsep{0pt}
\item Despite an overall productivity/performance trade-off,
more tightly-knit WikiProjects tend to produce articles more quickly as well as produce higher-quality articles;
\item The high performance of tightly-knit networks is consistent with individuals showing susceptibility to social influence;
\item Unequal participation is associated with lower performance;
\item The agent-based model shows that the relationship between quality and productivity can be influenced by both network structure and individual behavior.
\end{itemize}

@@

>>>>

The ability of large groups to reach mutually agreeable decisions is key to
democratic governance, organizational effectiveness, social movements, and peer production.
Faced with the intractability of large-scale participatory decision-making, traditional
systems have sacrificed one or more desirable properties, such as participation
(as in representative democracy), deliberation (as in voting), equality
(as in command hierarchies), and speed (as in formal consensus).
This dissertation examines the emerging role of internet-enabled collaborative
networks in overcoming these historical limitations.

In recent years, examples of large-scale collaborations have emerged that seem
to achieve the previously unachievable. Millions of volunteer Wikipedia editors
have created a high-quality encyclopedia without formal centralized leadership
\cite{keegan_evolution_2017, giles_internet_2005}.
The Free Software movement has produced the Linux kernel and GNU operating
system, which power much of the modern internet
\cite{coleman_coding_2012, benkler_coases_2002, raymond_cathedral_1999}.
Social movements such as the Arab Spring, Occupy, Black Lives Matter, and
Podemos have brought attention to deeply entrenched social issues,
and in some cases, contriubted to political regime changes
\cite{tufekci_twitter_2017, gonzalez-bailon_networked_2016}.
Participatory governance at such scales is unprecedented.
The emergence of these large-scale decentralized collaborations has been
attributed to the fast, bidirectional,
and global communication enabled by the internet
\cite{tufekci_twitter_2017, benkler_coases_2002}.
A better understanding of how specifically such communication sidesteps
historical barriers to large-scale collaboration will contribute to more
effective policy as well as best practices for organizational design and
intervention.
This dissertation focuses on one particularly challenging aspect of
such collaborations: decision-making.
Specifically, I examine how
when groups are too large for all members to participate in all discussions,
the course and outcome of the decision-making process is influenced by the
communication network structure: the shape of who talks to whom.

In this dissertation, I combine observational, agent-based modeling, and experimental
methods to evaluate the role of network structure in large-scale participatory decision-making.
I focus on group sizes well above the estimated 5--8 individual maximum for effective small goups
\cite{freeman_tyranny_1972, lohman_designing_2000, miflin_small_2004},
ranging from 32--33 participants in the experimental study to millions of editors in the
observational study of English-language Wikipedia WikiProjects.
Taken together, the findings in these studies suggests that small, tightly-knit pods,
interlocked through the sharing of members can benefit such collaborations.
This common structure, which I refer to as network deliberation, appears to preserve some
benefits of small-scale deliberation, while allowing efficient diffusion of information
through large groups.
Furthermore, I find evidence that network deliberation can mitigate some of the negative
aspects of social influence.
These findings provide insight into the success of existing large-scale collaborations
as well as provide guidance for the design of new sociotechnical systems.

@@

The fundamental challenge of large-scale collective decision-making is how to
reconcile the conflicting preferences of individual group members.
This challenge has been studied formally in social choice theory,
a sub-field of economics.
Prior work in social choice theory has found, somewhat discouragingly,
that even when all individual preferences are known perfectly,
making a fair collective decision isn't always possible.
In some cases, the method of aggregating individual preferences
(i.e. the voting system) can influence the outcome.
Social choice theory focuses on understanding of these limitations,
such as the Condorcet Paradox
\cite{condorcet_essay_1785} and
Arrow's Impossibility Theorem \cite{arrow_social_2012}.
While social choice theory typically assumes fixed fundamental preferences,
preferences can be instrumental in nature,
varying with factors such as available knowledge, group identity, or perspective.
This dissertation finds hope in the transformative potential of the
deliberative process.
Social choice theory typically assumes fixed individual preferences.
Deliberation allows individuals to influence and change each other's
preferences,
which creates the potential to sidestep the historical limitations of
social choice theory.
When individual preferences are allowed to vary, it becomes possible for an
irreconcilable set of preferences to evolve into one with a clear winner.
So far, this possibility has received relatively little attention,
most likely due to the historical intractability of large-scale deliberative
decision-making.
This dissertation explores the potential of effective large-scale deliberation, as enabled by the internet.
Such large-scale deliberation creates the potential for members to
resolve conflicting preferences and reach mutually acceptable decisions
without relying on coercive or hierarchical processes that might introduce
power imbalances or informational biases.

@@

Throughout this dissertation,
I motivate and develop a novel theoretical framework,
which I call {\em network deliberation}.
Network deliberation is an analytical framework that can be applied both to the empirical observation of existing sociotechical systems,
as well as to guide the creation and analysis of simplified models.
My review of the literature identifies a common theme
among successful large-scale internet-enabled collaborations:
large collectives composed of interlocking smaller groups.
These groups have various names, including:
committees, working groups, teams, circles, cores, syndicates,
affinity groups, zones, and nodes.
As an abstraction of these small interlocking group, I will use the term ``pods.''
Network deliberation describes large-scale collective deliberation achieved
through interlocking pods.
As in the theories of interlocking directorates \cite{levine_study_1979},
interlocking publics \cite{habermas_structural_1991},
and network rotation \cite{salehi_hive_2018},
pods allow for beneficial small group dynamics,
while the overlap between pods enables diffusion of information and opinions
through the greater collective.
The network deliberation framework abstracts and generalizes phenomena
such as interlocking directorates and interlocking publics,
allowing their commonalities to be studied from the perspective of network structure.
The framework also provides concrete parameters such as pod size and pod assignment method,
which can be used in quantitative analysis of models and experiments,
and provides a common language for making comparisons across models and experiments.
Network deliberation is conceptually similar to network rotation in that both describe
large-scale deliberation through repeated small-group interactions.
There are two key differences, one conceptual, and one relating to network structure.
Conceptually, network rotation reassigns pod membership one individual at a time in order
to optimize certain network properties,
while network deliberation focuses on providing an analytical framework for understanding
the impact of different pod assignment methods on network structure and deliberation outcome.
The one-at-a-time reassignments of network rotation also creates pods with high overlap,
while network deliberation is focused on simultaneous or sequential pods with little overlap.

In network deliberation, the method of assigning individuals to pods
(whether deliberate or self-organized) can produce interpersonal networks with
varying structures.
The central questions of this dissertation are 1. does an interlocking-pod structure improve deliberation relative to conventional structures? and 2. how does the structure of these interlocking-pod networks influence the process and outcome of deliberation
in large collaborations?

@@

\section{Contributions}
This dissertation describes the contributions of three projects.
Chapter \ref{chap:wp-prod-perf} describes an observational and computational
study of WikiProjects on the English-language Wikipedia.
This study examines one of the real-world collaborations that inspired the network deliberation
framework and shows that variations in the network structure of such a collaboration can
correlate with quality indicators.
Specifically, Chapter \ref{chap:wp-prod-perf} reports the following contributions:
\begin{itemize}
\setlength\itemsep{0pt}
\item Despite an overall productivity/performance trade-off,
more tightly-knit WikiProjects tend to produce articles more quickly as well as produce higher-quality articles;
\item The high performance of tightly-knit networks is consistent with individuals showing susceptibility to social influence;
\item Unequal participation is associated with lower performance;
\item The agent-based model shows that the relationship between quality and productivity can be influenced by both network structure and individual behavior.
\end{itemize}

@@

The above contributions suggest that network deliberation structure may be an important factor in the
success of existing large-scale, internet-enabled collaborations, and furthermore,
that it may help guide the design and organization of future collaborations.
Specifically, network deliberation appears to help moderate the effects of social influence,
providing some protection against both information cascades due to positive social influence,
and polarization due to negative social influence.
In summary, the work presented in the following chapters provides a framework for the
future study of large-scale participatory governance,
and provides evidence that large-scale participatory governance can be both practical and effective.

### 2022-08-09 UMich/front_materials/abstract.tex
Dissertation revisions to abstract and ch01.

As groups grow in size, they gain access to additional resources, creating opportunities for collective intelligence and collective action.
However, at very large scales, group decision-making becomes prohibitively slow and difficult to coordinate.
Traditional solutions include representative decision-making and/or a shift from deliberation to voting.
Both approaches sacrifice desirable properties.
Representative decision-making loses the potential benefits of collective intelligence and introduces hierarchies that may place the interests of specific individuals ahead of the interests of the group.
Voting sacrifices generativity: allowing a choice between predefined options, without allowing for improvement to those options.
Arrow’s impossibility theorem also fundamentally limits the fairness of voting.
This project proposes Networked Deliberation as a potential means of effective large-scale decision-making.
In Networked Deliberation, members of a large group are repeatedly partitioned into small deliberative pods.
Overlap between pods at different stages enables group-wide diffusion of information and preferences.
Different methods for assigning members to pods result in different network topologies.
This work combines observational study, agent-based modeling, and experiment to evaluate and better understand network deliberation.
An empirical observation of WikiProjects on the English-language Wikipeida identifies the network properties of the most effective projects.
A simple agent-based model of Network Deliberation shows improvements over conventional deliberation in the presence of strong social influence.
Finally, a controlled experiment studies network deliberation in a real world setting, tracking how individual preferences evolve, and finding evidence that network deliberation provides protection against negative consequences of social influence.
This work seeks to improve the ability of very large groups to quickly develop a consensus, enabling a more effective use of shared resources to achieve common goals.

>>>>

As groups grow in size, they gain access to additional resources, creating opportunities for collective intelligence and collective action.
However, at very large scales, group decision-making becomes prohibitively slow and difficult to coordinate.
Traditional solutions include representative decision-making and/or a shift from deliberation to voting.
Both approaches sacrifice desirable properties.
Representative decision-making loses the potential benefits of collective intelligence and introduces hierarchies that may place the interests of specific individuals ahead of the interests of the group.
Voting sacrifices generativity: allowing a choice between predefined options, without allowing for improvement to those options.
Arrow’s impossibility theorem also fundamentally limits the fairness of voting.
This project proposes Networked Deliberation as a potential means of effective large-scale decision-making.
In Networked Deliberation, members of a large group are repeatedly partitioned into small deliberative pods.
Overlap between pods at different stages enables group-wide diffusion of information and preferences.
Different methods for assigning members to pods result in different network topologies.
This work combines observational study, agent-based modeling, and an experiment to evaluate and better understand network deliberation.
An empirical observation of WikiProjects on the English-language Wikipedia identifies the network properties of the most effective projects.
A simple agent-based model of Network Deliberation shows improvements over conventional deliberation in the presence of strong social influence.
Finally, a controlled experiment studies network deliberation in a real world setting, tracking how individual preferences evolve, and finding evidence that network deliberation provides protection against negative consequences of social influence.
This work seeks to provide a framework that enables an understanding of
of how very large groups can quickly develop a consensus, enabling a more effective use of shared resources to achieve common goals.
